Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-10-25 numerics_paper/version2/numerical_chj.tex
Remove cyan coloring.

\subsection{Dam break problem on a dry bed}\label{sec:rp_dry_bed}
We start with the one-dimensional dam break problem on a dry bed.
This problem is a canonical example that demonstrates the `entropy glitch' of Roe's solver
at transonic rarefactions. Since the baseline Roe solver we use (see \S\ref{sec:roe}) does not contain
an entropy fix, it is important to demonstrate that the blended Riemann solver fixes the entropy glitch.
We follow the setup in \cite[\S 4.1.2]{delestre2013swashes}.
The domain is given by $x\in(0,10)$, and the initial condition is ${\color{cyan}u(x,0)=0}$ and
\begin{align*}
  h(x,0)=
  \begin{cases}
    h_l & x\leq x_0=5 \\
    h_r & x >x_0,
  \end{cases}
\end{align*}
with $h_l=5\times 10^{-3}$.  In our experiments we use $h_r=1\times
10^{-15}$ to avoid division by zero.
If we use Rusanov's solver with the first-order method \eqref{first-order_via_fluct}, 
the water height is guaranteed to remain positive. Using Roe or the blended solvers 
can introduce violations of positivity. Alternatively, if we consider the second-order scheme 
\eqref{second-order_via_fluct}, the water height could become negative regardless of the Riemann solver.
To reduce the occurance of loss of positivity, we only use the first-order method 
\eqref{first-order_via_fluct} for this test problem. 
Although using Roe or the blended solvers might still lead to loss of positivity, we 
did not observe that in our experiments.
The exact solution, which can also be found in \cite{delestre2013swashes} and references therein, is
\begin{align*}
  h(x,t) =
  \begin{cases}
    h_l, \\
    \frac{4}{9g}\left(\sqrt{gh_l}-\frac{x-x_0}{2t}\right)^2, \\
    0,
  \end{cases}
\quad
  u(x,t) =
  \begin{cases}
    0, &\mbox{ if } x\leq x_A(t), \\
    \frac{2}{3}\left(\sqrt{gh_l}+\frac{x-x_0}{t}\right), & \mbox{ if } x_A(t) < x\leq x_B(t), \\
    0, &\mbox{ if } x_B(t) < x,
  \end{cases}
\end{align*}
with $x_A(t)=x_0 - t\sqrt{gh_l}$ and $x_B=x_0+2t\sqrt{gh_l}$.
We solve the problem up to $t=10$.
In Figures \ref{fig:rp_dry_bed_roe}-\ref{fig:rp_dry_bed_blended},
we show the solution using method \eqref{first-order_via_fluct}
with Roe's solver, Rusanov's solver and the blended solver, respectively.
The entropy glitch, which is manifested as a non-physical shock at $x=x_0$,
is present with Roe's solver. Using Rusanov's and the blended solvers
fixes the entropy glitch.
In Table \ref{table:rp_dry_bed}, we summarize the results of a convergence test.
Due to the entropy glitch, we do not converge to the right solution with Roe solver,
hence the drop in rate of convergence. Using any of the other solvers, we converge to the 
correct solution. Note that using the blended solver leads to more accurate results. 
Since the method we use is first-order accurate, we get no better than first-order convergence.

@@

\subsection{Dam break problem on a wet bed}\label{sec:rp_wet_bed}
We consider now a one-dimensional dam break problem on a wet domain.
We follow the setup in \cite[\S 4.1.1]{delestre2013swashes}.
The domain is $x\in(0,10)$ and the initial condition is given by
${\color{cyan} u(x,0)=0}$ and
\begin{align*}
  h(x,0) =
  \begin{cases}
    h_l & x \leq x_0 \\
    h_r & x > x_0
  \end{cases}
\end{align*}
with $x_0=5$, $h_l=0.005$, and $h_r=0.001$.  The exact solution, which can be found in
\cite{delestre2013swashes} and references therein, is given by
\begin{align*}
  h(x,t) =
  \begin{cases}
    h_l, \\
    \frac{4}{9g}\left(\sqrt{gh_l}-\frac{x-x_0}{2t}\right)^2, \\
    \frac{c_m^2}{g}, \\
    h_r,
  \end{cases}
\quad
  u(x,t) =
  \begin{cases}
    0, &\mbox{ if } x\leq x_A(t), \\
    \frac{2}{3}\left(\sqrt{gh_l}+\frac{x-x_0}{t}\right), & \mbox{ if } x_A(t) < x\leq x_B(t), \\
    2(\sqrt{gh_l}-c_m), & \mbox{ if } x_B(t)<x\leq x_C(t), \\
    0, &\mbox{ if } x < x_C(t),
  \end{cases}
\end{align*}
where $x_A(t)=x_0-t\sqrt{gh_l}$, $x_B(t)=x_0+t\left(2\sqrt{gh_l}-3c_m\right)$,
$x_C(t)=x_0+t\frac{2c_m^2\left(\sqrt{gh_l}-c_m\right)}{c_m^2-gh_r}$ and
$c_m$ is the solution of
$-8gh_rc_m^2\left(\sqrt{gh_l}-c_m\right)^2+\left(c_m^2-gh_r\right)^2\left(c_m^2+gh_r\right)=0$.
We solve the problem up to the final time $t=5$ using the
second-order method \eqref{second-order_via_fluct}
with Roe's, Rusanov's and the blended solvers. The solution with different refinement levels
and each Riemann solver is shown in Figure \ref{fig:rp_wet_bed}.
In Table \ref{table:rp_wet_bed}, we summarize the results of a convergence test.
Since the solution is non-smooth, we expect no more than first order convergence rates. Note that the
results with the entropy dissipative blended solver are comparable to the results using Roe's solver.
That is, imposing entropy dissipation via the blended Riemann solver does not degrade the high-order
accuracy properties of Roe's solver.
In contrast, the accuracy and convergence rates using Rusanov's solver are clearly degraded.

@@

\subsection{Semi-analytical steady solution under rotational symmetry}\label{sec:steady_chj}
In this section, we consider the initial boundary value problem consisting of the
shallow water model \eqref{eq:sw} in an annular domain
\begin{align} \label{eq:annulus}
r_\textup{jet} \le r \le r_\infty
\end{align}
where $r = \sqrt{x^2+y^2}$, with prescribed inflow at $r=r_\textup{jet}$
and prescribed outflow at $r=r_\infty$.
The domain and boundary conditions are rotationally symmetric.
By assuming rotational symmetry in \eqref{eq:sw}, one obtains the system
\begin{subequations} \label{eq:rsw}
\begin{align}
    (rh)_t + (rhu)_r &= 0, \label{mass1} \\
    (rhu)_t + (rhu^2)_r + r \left(\frac{1}{2}gh^2\right)_r &= 0, \label{mom1}
\end{align}
\end{subequations}
where the depth $h$ and radial velocity $u$ are functions of $r$ and $t$.
By direct integration one finds that steady-state solutions of
\eqref{eq:rsw} satisfy
\begin{subequations}\label{steady}
\begin{align}
    rhu & = C \\
    h'(r) & = \frac{h(r)}{\frac{g}{{\color{cyan} C}^2} r^3 (h(r))^3 -r} = \frac{h(r)}{r} \cdot \frac{(F(r))^2}{1-(F(r))^2} \label{eq:dh}
\end{align}
\end{subequations}
for some $C$ independent of $r$.  Here $F(r)=|u(r)|/\sqrt{gh(r)}$ is the Froude number.
We see that two types of steady profiles exist, depending on whether the flow
is subcritical ($|F|<1$) or supercritical ($|F|>1$).  No smooth steady solution can
include both regimes, since the right hand side of \eqref{eq:dh} blows up when $F=1$.

\subsection{Numerical test: steady outflow}\label{sec:steady_outflow}
We now test the numerical methods by using a time-dependent simulation
to compute the steady flow solution just described, 
in the annulus $r\in(0.1,1)$ with constant inflow at $r=0.1$ and outflow at $r=1$.
The initial condition is
$h(r,t=0)=0.1$, ${\color{cyan}u(r,t=0)=0}$, the inner boundary condition is $h(0.1,t)=0.3$, ${\color{cyan}u(0.1,t)=0.75}$,
and the outer boundary condition is set to outflow
(see \cite[\S 21.8.5]{leveque2002finite} for details).
The computational mesh is logically quadrilateral, of the type shown in
Figure \ref{fig:mesh_chj}.
\begin{figure}[!h]
  \centering 
  \includegraphics[scale=0.15]{figures/mesh.png}
  \caption{
    Example of a grid for the problem with the two-dimensional CHJ.
    \label{fig:mesh_chj}}
\end{figure}
Regardless of the initial condition, the exact solution converges to a
steady state profile given by one of the two solutions of \eqref{steady},
corresponding to subcritical or supercritical flow.  In the present case we
have imposed a supercritical inflow.
In Figure \ref{fig:steady_outflow}, we show the solution and $\theta_i$ at different times using
the second-order method \eqref{second-order_via_fluct}
with Roe's, Rusanov's and the blended Riemann solvers.  
Additionally, in Table \ref{table:steady_outflow}, we summarize the results of a convergence study
based on methods \eqref{first-order_via_fluct} and
\eqref{second-order_via_fluct}, using the same Riemann solvers.
Although the chosen initial
condition leads initially to shock formation, the steady state is smooth and close to
second order convergence is observed for the Roe and blended solvers.

>>>>

\subsection{Dam break problem on a dry bed}\label{sec:rp_dry_bed}
We start with the one-dimensional dam break problem on a dry bed.
This problem is a canonical example that demonstrates the `entropy glitch' of Roe's solver
at transonic rarefactions. Since the baseline Roe solver we use (see \S\ref{sec:roe}) does not contain
an entropy fix, it is important to demonstrate that the blended Riemann solver fixes the entropy glitch.
We follow the setup in \cite[\S 4.1.2]{delestre2013swashes}.
The domain is given by $x\in(0,10)$, and the initial condition is $u(x,0)=0$ and
\begin{align*}
  h(x,0)=
  \begin{cases}
    h_l & x\leq x_0=5 \\
    h_r & x >x_0,
  \end{cases}
\end{align*}
with $h_l=5\times 10^{-3}$.  In our experiments we use $h_r=1\times
10^{-15}$ to avoid division by zero.
If we use Rusanov's solver with the first-order method \eqref{first-order_via_fluct}, 
the water height is guaranteed to remain positive. Using Roe or the blended solvers 
can introduce violations of positivity. Alternatively, if we consider the second-order scheme 
\eqref{second-order_via_fluct}, the water height could become negative regardless of the Riemann solver.
To reduce the occurance of loss of positivity, we only use the first-order method 
\eqref{first-order_via_fluct} for this test problem. 
Although using Roe or the blended solvers might still lead to loss of positivity, we 
did not observe that in our experiments.
The exact solution, which can also be found in \cite{delestre2013swashes} and references therein, is
\begin{align*}
  h(x,t) =
  \begin{cases}
    h_l, \\
    \frac{4}{9g}\left(\sqrt{gh_l}-\frac{x-x_0}{2t}\right)^2, \\
    0,
  \end{cases}
\quad
  u(x,t) =
  \begin{cases}
    0, &\mbox{ if } x\leq x_A(t), \\
    \frac{2}{3}\left(\sqrt{gh_l}+\frac{x-x_0}{t}\right), & \mbox{ if } x_A(t) < x\leq x_B(t), \\
    0, &\mbox{ if } x_B(t) < x,
  \end{cases}
\end{align*}
with $x_A(t)=x_0 - t\sqrt{gh_l}$ and $x_B=x_0+2t\sqrt{gh_l}$.
We solve the problem up to $t=10$.
In Figures \ref{fig:rp_dry_bed_roe}-\ref{fig:rp_dry_bed_blended},
we show the solution using method \eqref{first-order_via_fluct}
with Roe's solver, Rusanov's solver and the blended solver, respectively.
The entropy glitch, which is manifested as a non-physical shock at $x=x_0$,
is present with Roe's solver. Using Rusanov's and the blended solvers
fixes the entropy glitch.
In Table \ref{table:rp_dry_bed}, we summarize the results of a convergence test.
Due to the entropy glitch, we do not converge to the right solution with Roe solver,
hence the drop in rate of convergence. Using any of the other solvers, we converge to the 
correct solution. Note that using the blended solver leads to more accurate results. 
Since the method we use is first-order accurate, we get no better than first-order convergence.

@@

\subsection{Dam break problem on a wet bed}\label{sec:rp_wet_bed}
We consider now a one-dimensional dam break problem on a wet domain.
We follow the setup in \cite[\S 4.1.1]{delestre2013swashes}.
The domain is $x\in(0,10)$ and the initial condition is given by
$u(x,0)=0$ and
\begin{align*}
  h(x,0) =
  \begin{cases}
    h_l & x \leq x_0 \\
    h_r & x > x_0
  \end{cases}
\end{align*}
with $x_0=5$, $h_l=0.005$, and $h_r=0.001$.  The exact solution, which can be found in
\cite{delestre2013swashes} and references therein, is given by
\begin{align*}
  h(x,t) =
  \begin{cases}
    h_l, \\
    \frac{4}{9g}\left(\sqrt{gh_l}-\frac{x-x_0}{2t}\right)^2, \\
    \frac{c_m^2}{g}, \\
    h_r,
  \end{cases}
\quad
  u(x,t) =
  \begin{cases}
    0, &\mbox{ if } x\leq x_A(t), \\
    \frac{2}{3}\left(\sqrt{gh_l}+\frac{x-x_0}{t}\right), & \mbox{ if } x_A(t) < x\leq x_B(t), \\
    2(\sqrt{gh_l}-c_m), & \mbox{ if } x_B(t)<x\leq x_C(t), \\
    0, &\mbox{ if } x < x_C(t),
  \end{cases}
\end{align*}
where $x_A(t)=x_0-t\sqrt{gh_l}$, $x_B(t)=x_0+t\left(2\sqrt{gh_l}-3c_m\right)$,
$x_C(t)=x_0+t\frac{2c_m^2\left(\sqrt{gh_l}-c_m\right)}{c_m^2-gh_r}$ and
$c_m$ is the solution of
$-8gh_rc_m^2\left(\sqrt{gh_l}-c_m\right)^2+\left(c_m^2-gh_r\right)^2\left(c_m^2+gh_r\right)=0$.
We solve the problem up to the final time $t=5$ using the
second-order method \eqref{second-order_via_fluct}
with Roe's, Rusanov's and the blended solvers. The solution with different refinement levels
and each Riemann solver is shown in Figure \ref{fig:rp_wet_bed}.
In Table \ref{table:rp_wet_bed}, we summarize the results of a convergence test.
Since the solution is non-smooth, we expect no more than first order convergence rates. Note that the
results with the entropy dissipative blended solver are comparable to the results using Roe's solver.
That is, imposing entropy dissipation via the blended Riemann solver does not degrade the high-order
accuracy properties of Roe's solver.
In contrast, the accuracy and convergence rates using Rusanov's solver are clearly degraded.

@@

\subsection{Semi-analytical steady solution under rotational symmetry}\label{sec:steady_chj}
In this section, we consider the initial boundary value problem consisting of the
shallow water model \eqref{eq:sw} in an annular domain
\begin{align} \label{eq:annulus}
r_\textup{jet} \le r \le r_\infty
\end{align}
where $r = \sqrt{x^2+y^2}$, with prescribed inflow at $r=r_\textup{jet}$
and prescribed outflow at $r=r_\infty$.
The domain and boundary conditions are rotationally symmetric.
By assuming rotational symmetry in \eqref{eq:sw}, one obtains the system
\begin{subequations} \label{eq:rsw}
\begin{align}
    (rh)_t + (rhu)_r &= 0, \label{mass1} \\
    (rhu)_t + (rhu^2)_r + r \left(\frac{1}{2}gh^2\right)_r &= 0, \label{mom1}
\end{align}
\end{subequations}
where the depth $h$ and radial velocity $u$ are functions of $r$ and $t$.
By direct integration one finds that steady-state solutions of
\eqref{eq:rsw} satisfy
\begin{subequations}\label{steady}
\begin{align}
    rhu & = C \\
    h'(r) & = \frac{h(r)}{\frac{g}{C^2} r^3 (h(r))^3 -r} = \frac{h(r)}{r} \cdot \frac{(F(r))^2}{1-(F(r))^2} \label{eq:dh}
\end{align}
\end{subequations}
for some $C$ independent of $r$.  Here $F(r)=|u(r)|/\sqrt{gh(r)}$ is the Froude number.
We see that two types of steady profiles exist, depending on whether the flow
is subcritical ($|F|<1$) or supercritical ($|F|>1$).  No smooth steady solution can
include both regimes, since the right hand side of \eqref{eq:dh} blows up when $F=1$.

\subsection{Numerical test: steady outflow}\label{sec:steady_outflow}
We now test the numerical methods by using a time-dependent simulation
to compute the steady flow solution just described, 
in the annulus $r\in(0.1,1)$ with constant inflow at $r=0.1$ and outflow at $r=1$.
The initial condition is
$h(r,t=0)=0.1$, $u(r,t=0)=0$, the inner boundary condition is $h(0.1,t)=0.3$, $u(0.1,t)=0.75$,
and the outer boundary condition is set to outflow
(see \cite[\S 21.8.5]{leveque2002finite} for details).
The computational mesh is logically quadrilateral, of the type shown in
Figure \ref{fig:mesh_chj}.
\begin{figure}[!h]
  \centering 
  \includegraphics[scale=0.15]{figures/mesh.png}
  \caption{
    Example of a grid for the problem with the two-dimensional CHJ.
    \label{fig:mesh_chj}}
\end{figure}
Regardless of the initial condition, the exact solution converges to a
steady state profile given by one of the two solutions of \eqref{steady},
corresponding to subcritical or supercritical flow.  In the present case we
have imposed a supercritical inflow.
In Figure \ref{fig:steady_outflow}, we show the solution and $\theta_i$ at different times using
the second-order method \eqref{second-order_via_fluct}
with Roe's, Rusanov's and the blended Riemann solvers.  
Additionally, in Table \ref{table:steady_outflow}, we summarize the results of a convergence study
based on methods \eqref{first-order_via_fluct} and
\eqref{second-order_via_fluct}, using the same Riemann solvers.
Although the chosen initial
condition leads initially to shock formation, the steady state is smooth and close to
second order convergence is observed for the Roe and blended solvers.
